---
title: "Introduction to GIS in R"
format: html
---

```{r, include=FALSE}

# load packages
library(sf) 
library(ggplot2)
library(dplyr)
library(readr)
library(stringr)
library(ggspatial)
library(plotly)
library(countdown)
library(tidyr)
```

## Aims

-   Know how to load spatial data into R using the `sf` library.
-   Be familiar with using GSS codes to join statistics to geographies.
-   Understand how spatial objects can be manipulated using `dplyr`.
-   Understand how to use spatial joins.
-   Be aware of map projections and Coordinate Reference Systems (CRS) and be able to modify them.
-   Know how to make static and interactive maps in `ggplot2`.
-   Be able to export your maps and shapefiles.

## GIS and R

R is commonly used for statistical analysis and programming, however it also has a range of geospatial libraries developed by a community of researchers and programmers. In the last few years, working with spatial data became much easier in R, with the development of the `sf` package. `sf` keeps all the spatial information for each observation in a geometry column which means that we can treat it like a normal data frame and also perform spatial operations on the data.

```{r, echo=FALSE}
poi <- sf::st_read("/Users/christophersteinberg/Downloads/hotosm_ken_points_of_interest_points_shp/hotosm_ken_points_of_interest_points_shp.shp", quiet=TRUE, as_tibble = TRUE)
head(poi)
```

## Working with spatial data in R

### Open Street Map Data

Throughout this tutorial you will be using data about Kenyan rainfall - [Kenyan Rainfall Data](https://data.humdata.org/dataset/ken-rainfall-subnational). It covers total rainfall at the subcounty level in the last 5 years, from 2021-01-01. The data is updated weekly and includes a range of variables capturing amount of rain.

We want to visualise, and better understand how much rain has fallen in the last 5 years at the sub-county and county level, and what the distribution is at the MSOA level of geography. To achieve this we will have to import spatial data, manipulate it, create summary statistics, and then plot it.

## Loading spatial and non-spatial data

Rainfall data has been tidied up and saved as a Comma Separated Value file (.csv). We can use `read_csv` to open it in R.

### Exercise

`r countdown(minutes = 10, seconds = 0, style = "position: relative; width: min-content;")`

1.  Create a new object called `poi` by using `read_csv()`. Load data located in "<https://data.humdata.org/dataset/ken-rainfall-subnational>".
2.  Use `glimpse()` or `head()` to view `health` structure.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>\

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r, eval = TRUE}
poi <- readr::read_csv("/Users/christophersteinberg/Documents/GitHub/intro_R-main/data/poi.csv")

poi
```

</details>

<br>

`poi` is currently just a data frame - it has not got an explicit geometry column which links observations to their geographic location. It does however contain several columns which can be used to convert it into a spatial data format.

**Ward_code** column references the GSS codes of wards within which the observations fall. GSS codes can be used to join `lfb` data to boundaries from the Open Geography Portal. One issue with this particular column is that it does not indicate the currency of GSS codes. Wards are subject to frequent change, and as such it is best practice to be clear about the dates of any boundaries used by stating the exact code used, e.g. **wd19cd**. Because LFB data does not include this information we have no guarantee that the boundaries and GSS codes we join will match.

Fortunately we have also been provided with columns recording the **easting**, and **northing** of each incident. We can use those to convert `lfb` into an `sf` object. To achieve this we will use the `st_as_sf()` function which takes the following arguments:

```{r, eval = FALSE}
new_object <- st_as_sf(x = input_data_frame, coords = c("x_coordinate_column", "y_coordinate_column"), crs = 27700)
```

### Exercise

`r countdown(minutes = 10, seconds = 0, style = "position: relative; width: min-content;")`

1.  Create a new object called `poi_sf` by converting `poi` using the `st_as_sf()` function.
2.  Use `glimpse()` or `head()` to view `poi_sf` structure.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r}
poi_sf <- sf::st_as_sf(x = poi, coords = c("easting", "northing"), crs = 4326)
head(poi_sf)
```

</details>

<br>

`st_as_sf()` converted the **easting** and **northing** columns to simple feature geometries and created a new column called **geometry** which holds spatial information for each row. Now that `poi` is a spatial object we can plot it using the `ggplot2` package. For now we will use the `geom_sf()` function which creates a quick map, using `ggplot2's` default settings. `geom_sf()` only needs to be supplied with a simple feature object and is very useful for quickly inspecting your data. To quickly plot multiple layers on the same map use `geom_sf() + geom_sf()`.

### Exercise

`r countdown(minutes = 3, seconds = 0, style = "position: relative; width: min-content;")`

1.  Plot `poi_sf` using the `geom_sf()` function.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r}
ggplot2::ggplot(poi_sf) + 
  ggplot2::geom_sf() 
```

</details>

<br>

We can also create interactive maps using `plotly` or `leaflet`. For `plotly` we ca write the map using ggplot2, before executing `ggplotly()`.

### Exercise

`r countdown(minutes = 3, seconds = 0, style = "position: relative; width: min-content;")`

1.  Make an interactive map of `poi_sf` using the `ggplotly()` function.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r}
ggplot_graph <- ggplot2::ggplot(poi_sf) + 
  ggplot2::geom_sf() 

plotly::ggplotly(ggplot_graph)
  
```

</details>

<br>

## Filtering by Administration Code

This dataset covers all of Kenya at subnational administrative boundaries. Let's look at solely locations in Nairobi to narrow down our exploration more. To remove all points outside of Nairobi we will have to import a shapefile with the right county level and then use it to spatially filter our `poi_df` data.

So far we have created our own `sf` objects by adding a geometry column. The kenya data set is already a spatial one and as such we can use the `st_read()` function from the `sf` package to import it. `st_read` isextremely versatile and able to import most spatial data formats into R. The only argument that needs to be supplied to `st_read` is the fullpath to the UTLA file

### Exercise

`r countdown(minutes = 10, seconds = 0, style = "position: relative; width: min-content;")`

1.  Use `st_read()` to load the kenya sub-county boundaries you downloaded at the beginning of the tutorial, as`kenya_2017`.
2.  UTLA path - `https://data.humdata.org/dataset/cod-ab-ken`
3.  Make a static map of the object you have just created using `geom_sf()`.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r}

kenya_2017 <- sf::st_read("/Users/christophersteinberg/Documents/GitHub/admn2/ken_admbnda_adm2_iebc_20191031.shp", quiet = TRUE)
ggplot2::ggplot(kenya_2017) + 
  ggplot2::geom_sf() 

```

</details>

<br>

Sub-county level boundaries have loaded correctly but they currently cover all of Kenya. Because simple feature objects are data frames with a geometry column attached, any operations that we would perform on a normal data frame can also be performed on an object of class `sf`. We will use `dplyr::filter` and `stringr::str_detect()` to only keep UTLAs where their `ADMN_PCODE` code starts with "KE47". "KE47" denotes that an sub-county is part of Nairobi county.

### Exercise

`r countdown(minutes = 10, seconds = 0, style = "position: relative; width: min-content;")`

1.  Inspect kenya_2017 using `head()` or `glimpse()`, and identify which column holds the GSS codes - it should end in "cd".
2.  Create a new object called `nairobi_subcounty`. Use `dplyr::filter` alongside `stringr::str_detect()` to only keep observations which have a Administrative code starting with "KE47".
3.  Plot `nairobi_subcounty` to see if the results look correct.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r}
head(kenya_2017)
```

```{r}
nairobi_subcounty <- dplyr::filter(kenya_2017, stringr::str_detect(ADM2_PCODE, "KE047"))
ggplot2::ggplot(nairobi_subcounty) + 
  ggplot2::geom_sf() 
```

</details>

<br>

Finally, for the next step, we only need the outer boundary of Nairobi - all the internal subcounty boundaries have to be removed and only the outer edges kept. `sf` has a function exactly for this purpose called `st_union()`. It only takes one argument, which is the `sf` object we want to merge.

### Exercise

`r countdown(minutes = 10, seconds = 0, style = "position: relative; width: min-content;")`

1.  Create a new object called `nairobi_boundary` using the `st_union` function.
2.  Plot it to check the results.

<details>

<summary style="font-size: 20px; font-weight: bold;">

5.2.2. Solution</strong>

</summary>

<br>

```{r}
nairobi_boundary <- sf::st_union(nairobi_subcounty)
ggplot2::ggplot(nairobi_boundary) + 
  ggplot2::geom_sf() 
```

</details>

<br>
